ActiveCA: Time Use Data from the General Social Survey of Canada to Study Active Travel

### 2025-05-30 paper/Manuscript-Data-Package-1Round-Review.Rmd
Changing the data paper due to the revision process.

# How to use {ActiveCA}

This section presents some potential applications of the {ActiveCA} R package. In fact, we expect that the application of this package to extend beyond our pre-imagined range of uses.

To start the demonstration, it is first necessary to install the package. The code chunk below demonstrates how to install the {ActiveCA} package in R:

```{r instal-ActiveCA, echo=TRUE}
if(!require("ActiveCA", character.only = TRUE)) {
      remotes::install_github("dias-bruno/ActiveCA")
}
```

After installing the {ActiveCA} R package, you can load it using the following code:

```{r load-ActiveCA, echo=TRUE}
library(ActiveCA)
```

The documentation of the package can be accessed by running `?ActiveCA` in `R`. The documentation includes a description of the package, information about the authors, contributors, and maintainers, and an overview of all pre-processed data sets included in the package.

Figure \ref{fig:walking_documentation} displays the documentation for the walking episodes from the 2022 GSS (`walking_2022.rda`). The documentation provides a brief description of the dataset, instructions on how to load the package, as well as details about the file format and descriptions of all variables (already labeled).

```{r walking_documentation, echo=FALSE, out.width="100%", fig.cap=paste0("Documentation for the dataset of walking episodes from 2022."), fig.align="center"}

knitr::include_graphics(paste0(here::here(),"/paper/Manuscript-figures/walking-2022-documentation.png"))
```

The code below shows how to load the walking episodes from the 2022 GSS (Cycle 29):

```{r load-walking-2022-episodes, echo=TRUE}
data(walking_2022)
```

The GSS surveys apply a probability sampling methodology, in which each episode or person selected in the sample represents several other episodes or persons not in the sample. The number of episodes and persons represented by a episode or person is determined by the weight or weighting factor. Because of this, estimates of the number of episodes or persons need to be calculated applying the corresponding weighting factors.

For instance, to calculate the percentage of respondents from the 2022 GSS survey with active travel episodes, it is necessary to account for the person's weight. In 2022, the weight variable is represented by `WGHT_PER.` The code below demonstrates how to obtain the percentage of people with active travel episodes by age group. It uses the `dplyr` package [@dplyr2023] to manipulate the data.

The process begins by creating a dataset that sums the population by age group. Then, it joins the 2022 episodes with the 2022 Main File. Note that in both operations, the code sums the person weight variable (`WGHT_PER`) to obtain the correct population values. After this, a new dataset, `Active_percentage`, is created by merging both previous datasets. The percentage of active travel episodes by age group is calculated by dividing the total population by the population with active trip episodes, then multiplying by 100 and rounding with 2 decimal places.

```
# Load the dplyr library for data manipulation
library('dplyr')

# Load the episodes dataset
data(gss_episodes)

# Calculate the total population for each age group
Total_population <- gss_main_2022 |> 
  group_by(AGEGR10) |> 
  summarise(Total_population  = sum(WGHT_PER))

# Calculate the active population for each age group
Active_population <- gss_episodes |> 
      filter(YEAR == 2022) |> 
  left_join(gss_main_2022, by = c("PUMFID"="PUMFID")) |> 
  group_by(AGEGR10) |> 
  summarise(Active_population  = sum(WGHT_PER))

# Combine total and active population 
#and calculate the percentage
Active_percentage <- Total_population |> 
  left_join(Active_population, by = c("AGEGR10" = "AGEGR10")) |>
  mutate(
    Percentage = round(Active_population/Total_population*100,2))
```
The result shows that the age group with the highest share of active trips is those between 15 and 20 years old, with almost 44%, followed by those between 25 and 34 years old with around 19%. There is a significant drop in percentage for the following groups, with the percentage falling to between 12% and 15% for the remaining age groups.

```{r percentages-act, echo=FALSE, out.width='80%'}
# Load the dplyr library for data manipulation
library('dplyr')

# Load the episodes dataset
data(gss_episodes)

# Calculate the total population for each age group
Total_population <- gss_main_2022 |> 
  group_by(AGEGR10) |> 
  summarise(Total_population  = sum(WGHT_PER))

# Calculate the active population for each age group
Active_population <- gss_episodes |> 
      filter(YEAR == 2022) |> 
  left_join(gss_main_2022, by = c("PUMFID"="PUMFID")) |> 
  group_by(AGEGR10) |> 
  summarise(Active_population  = sum(WGHT_PER))

# Combine total and active population 
#and calculate the percentage
Active_percentage <- Total_population |> 
  left_join(Active_population, by = c("AGEGR10" = "AGEGR10")) |>
  mutate(
    Percentage = round(Active_population/Total_population*100,2))

# Display the final result
Active_percentage |>
  kable()
```

## Processed data set

@@

{ActiveCA} also enables obtaining insights from the main processed files. Figure \ref{fig:figure-stress} present how the level of stress varied among respondents depending on their marital status in 2015. According to this plot, married respondents reported the highest level of stress, relating to feel stressed every day, with 17% of possible cases.

```{r main_stress_figure, include = FALSE}
main_stress_figure <- gss_main_2015 |>
  filter(!is.na(MARSTAT), !is.na(GTU_110)) |>
  group_by(MARSTAT, GTU_110) |>
  dplyr::summarise(n = sum(WGHT_PER), .groups = "drop") |>
  mutate(percentage = n / sum(n) * 100) |>
  ungroup() |>
  ggplot(aes(x = MARSTAT,
    y = GTU_110)) +
    geom_tile(aes(fill = percentage)) +
    labs(x = "Marital status of the respondent",
    y = "How often do you feel rushed?",
    fill = "Percentage (%)") + 
    #geom_text(aes(label = round(percentage)), color = "black", size = 3) + 
    theme_bw() + 
    scale_fill_gradientn(colors = RColorBrewer::brewer.pal(9, "YlGn")) +
    theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1))

# Save the combined plot to a file
ggsave(file = paste0(here::here(),"/paper/Manuscript-figures/main_stress_figure.jpg"), 
       plot = main_stress_figure,
       width = 12, 
       height = 8, 
       units = "cm", 
       dpi = 300)
```

>>>>

# {ActiveCA} data sets

This section presents some potential applications of the {ActiveCA} R package. In fact, we expect that the application of this package to extend beyond our pre-imagined range of uses. The installation instruction and also some examples of application of the {ActiveCA} R package are available in the `vignettes`, available in the Github repository.

## Active episodes

@@

{ActiveCA} also enables obtaining insights from the main processed files. Figure \ref{fig:figure-stress} present how the level of stress varied among respondents depending on their marital status in 2022. According to this plot, married respondents reported the highest level of stress, relating to feel stressed every day, with 15% of possible.

```{r main_stress_figure, include = FALSE}
main_stress_figure <- gss_main_2022 |>
  filter(!is.na(MARSTAT), !is.na(GTU_110)) |>
  group_by(MARSTAT, GTU_110) |>
  dplyr::summarise(n = sum(WGHT_PER), .groups = "drop") |>
  mutate(percentage = n / sum(n) * 100) |>
  ungroup() |>
  ggplot(aes(x = GTU_110,
    y = MARSTAT)) +
    geom_tile(aes(fill = percentage)) +
    labs(x = "Marital status of the respondent",
    y = "How often do you feel rushed?",
    fill = "Percentage (%)") + 
    geom_text(aes(label = round(percentage)), color = "black", size = 3) + 
    theme_bw() + 
    scale_fill_gradientn(colors = RColorBrewer::brewer.pal(9, "YlGn")) +
    theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1))

# Save the combined plot to a file
ggsave(file = paste0(here::here(),"/paper/Manuscript-figures/main_stress_figure.jpg"), 
       plot = main_stress_figure,
       width = 15, 
       height = 10, 
       units = "cm", 
       dpi = 300)
```

### 2025-05-30 paper/Manuscript-Data-Package-1Round-Review.tex
Changing the data paper due to the revision process.

% From pandoc table feature
\usepackage{longtable,booktabs,array}
\usepackage{calc} % for calculating minipage widths
% Correct order of tables after \paragraph or \subparagraph
\usepackage{etoolbox}
\makeatletter
\patchcmd\longtable{\par}{\if@noskipsec\mbox{}\fi\par}{}{}
\makeatother
% Allow footnotes in longtable head/foot
\IfFileExists{footnotehyper.sty}{\usepackage{footnotehyper}}{\usepackage{footnote}}
\makesavenoteenv{longtable}

@@

\section{How to use \{ActiveCA\}}\label{how-to-use-activeca}

This section presents some potential applications of the \{ActiveCA\} R
package. In fact, we expect that the application of this package to
extend beyond our pre-imagined range of uses.

To start the demonstration, it is first necessary to install the
package. The code chunk below demonstrates how to install the
\{ActiveCA\} package in R:

\begin{verbatim}
if(!require("ActiveCA", character.only = TRUE)) {
      remotes::install_github("dias-bruno/ActiveCA")
}
\end{verbatim}

After installing the \{ActiveCA\} R package, you can load it using the
following code:

\begin{verbatim}
library(ActiveCA)
\end{verbatim}

The documentation of the package can be accessed by running
\texttt{?ActiveCA} in \texttt{R}. The documentation includes a
description of the package, information about the authors, contributors,
and maintainers, and an overview of all pre-processed data sets included
in the package.

Figure \ref{fig:walking_documentation} displays the documentation for
the walking episodes from the 2022 GSS (\texttt{walking\_2022.rda}). The
documentation provides a brief description of the dataset, instructions
on how to load the package, as well as details about the file format and
descriptions of all variables (already labeled).

{\centering \includegraphics[width=1\linewidth]{Manuscript-figures/walking-2022-documentation}

\caption{Documentation for the dataset of walking episodes from 2022.}\label{fig:walking_documentation}
\end{figure}

The code below shows how to load the walking episodes from the 2022 GSS
(Cycle 29):

\begin{verbatim}
data(walking_2022)
\end{verbatim}

The GSS surveys apply a probability sampling methodology, in which each
episode or person selected in the sample represents several other
episodes or persons not in the sample. The number of episodes and
persons represented by a episode or person is determined by the weight
or weighting factor. Because of this, estimates of the number of
episodes or persons need to be calculated applying the corresponding
weighting factors.

For instance, to calculate the percentage of respondents from the 2022
GSS survey with active travel episodes, it is necessary to account for
the person's weight. In 2022, the weight variable is represented by
\texttt{WGHT\_PER.} The code below demonstrates how to obtain the
percentage of people with active travel episodes by age group. It uses
the \texttt{dplyr} package \citep{dplyr2023} to manipulate the data.

The process begins by creating a dataset that sums the population by age
group. Then, it joins the 2022 episodes with the 2022 Main File. Note
that in both operations, the code sums the person weight variable
(\texttt{WGHT\_PER}) to obtain the correct population values. After
this, a new dataset, \texttt{Active\_percentage}, is created by merging
both previous datasets. The percentage of active travel episodes by age
group is calculated by dividing the total population by the population
with active trip episodes, then multiplying by 100 and rounding with 2
decimal places.

\begin{verbatim}
# Load the dplyr library for data manipulation
library('dplyr')

# Load the episodes dataset
data(gss_episodes)

# Calculate the total population for each age group
Total_population <- gss_main_2022 |> 
  group_by(AGEGR10) |> 
  summarise(Total_population  = sum(WGHT_PER))

# Calculate the active population for each age group
Active_population <- gss_episodes |> 
      filter(YEAR == 2022) |> 
  left_join(gss_main_2022, by = c("PUMFID"="PUMFID")) |> 
  group_by(AGEGR10) |> 
  summarise(Active_population  = sum(WGHT_PER))

# Combine total and active population 
#and calculate the percentage
Active_percentage <- Total_population |> 
  left_join(Active_population, by = c("AGEGR10" = "AGEGR10")) |>
  mutate(
    Percentage = round(Active_population/Total_population*100,2))
\end{verbatim}

The result shows that the age group with the highest share of active
trips is those between 15 and 20 years old, with almost 44\%, followed
by those between 25 and 34 years old with around 19\%. There is a
significant drop in percentage for the following groups, with the
percentage falling to between 12\% and 15\% for the remaining age
groups.

\begin{longtable}[]{@{}lrrr@{}}
\toprule\noalign{}
AGEGR10 & Total\_population & Active\_population & Percentage \\
\midrule\noalign{}
\endhead
\bottomrule\noalign{}
\endlastfoot
15 to 24 years & 4537139 & 1984658.8 & 43.74 \\
25 to 34 years & 5364748 & 1048653.8 & 19.55 \\
35 to 44 years & 5227988 & 669688.8 & 12.81 \\
45 to 54 years & 4720990 & 605162.1 & 12.82 \\
55 to 64 years & 5143935 & 690336.5 & 13.42 \\
65 to 74 years & 4160941 & 619732.8 & 14.89 \\
75 years and over & 2981061 & 422799.5 & 14.18 \\
\end{longtable}

\subsection{Processed data set}\label{processed-data-set}

>>>>

@@

\section{\{ActiveCA\} data sets}\label{activeca-data-sets}

This section presents some potential applications of the \{ActiveCA\} R
package. In fact, we expect that the application of this package to
extend beyond our pre-imagined range of uses. The installation
instruction and also some examples of application of the \{ActiveCA\} R
package are available in the \texttt{vignettes}, available in the Github
repository.

\subsection{Active episodes}\label{active-episodes}
